---
title: "ShinyState Tutorial"
subtitle: "React-style components and hooks for Shiny"
author: "ShinyState"
date: "`r format(Sys.Date())`"
output:
  revealjs::revealjs_presentation:
    theme: moon
    transition: slide
    highlight: pygments
    center: false
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

## Welcome

**ShinyState** brings a React-inspired programming model to Shiny:

- Components with local state
- Hooks: `useState()`, `useEffect()`, `useMemo()`, `useReducer()`
- No hand-written `reactive()` / `observe()` in app code

---

## The problem

Classic Shiny spreads logic across:

- `reactive()` values
- `observe()` / `observeEvent()`
- `renderUI()` / `renderText()`

As apps grow, dependencies become hard to follow.

---

## The ShinyState idea

Think in **components**:

```r
component(
  id = "counter",
  state = useState(count = 0L),
  effect(deps = "count", function(state) { ... }),
  render = function(state, ns) {
    # hooks + UI
  }
)
```

Each component owns its state, effects, and render function.

---

## Install

```{r}
# install.packages("devtools")
devtools::install_local("path/to/ShinyState")

library(shiny)
library(ShinyState)
```

---

## Core workflow

1. Define components with `component()`
2. Mount UI with `mount()`
3. Run server logic with `serve()`

```{r}
shinyApp(
  ui = fluidPage(mount(counter)),
  server = function(input, output, session) {
    serve(counter, input, output, session)
  }
)
```

---

## `component()`

```{r}
counter <- component(
  id = "counter",
  state = useState(count = 0L),
  effect(
    deps = "count",
    function(state) {
      message("Count is now: ", state$count)
    }
  ),
  render = function(state, ns) {
  }
)
```

Returns a list with `ui`, `server`, and metadata.

---

## `useState()`

Declare initial state at component definition time:

```{r}
state = useState(
  page = 1L,
  filter = NULL
)
```

Inside `render()`, read and update via the `state` accessor:

```{r}
state$count          # read
state$set(count = 5) # write + re-render
state$all()          # snapshot
```

---

## State updates

Functional updates are supported:

```{r}
state$update(function(current) {
  list(count = current$count + 1L)
})
```

`state$set(...)` triggers a component re-render automatically.

---

## `effect()` / `useEffect()`

Run side effects when dependencies change.

**Declarative** (top-level in `component()`):

```{r}
effect(
  deps = "filter",
  function(state) {
    state$set(page = 1L)  # reset page when filter changes
  }
)
```

- `deps = NULL` → run once on mount
- Return a function from the effect to run cleanup

---

## `useMemo()`

Cache expensive derived values:

```{r}
render = function(state, ns) {
  doubled <- useMemo(
    function() state$count * 2L,
    deps = "count"
  )

  p(paste("Doubled:", doubled))
}
```

Recomputes only when `count` changes.

---

## `useReducer()`

For state transitions with actions:

```{r}
render = function(state, ns) {
  reducer <- useReducer(
    function(prev, action) {
      switch(action,
        add = c(prev, "new"),
        drop = if (length(prev)) prev[-length(prev)] else prev,
        prev
      )
    },
    initial = character()
  )

  reducer$dispatch("add")
  reducer$state
}
```

---

## `useCallback()` + `bindButton()`

Wire button clicks to state updates:

```{r}
render = function(state, ns) {
  useCallback("inc", function(s) {
    s$set(count = s$count + 1L)
  })

  bindButton(ns, "inc", "+")
}
```

**Important:** use `bindButton()`, not `actionButton()`, inside `render()`.

---

## Why bound helpers?

`render()` output is rebuilt when state changes.

Plain Shiny inputs and buttons reset on re-render and can fire handlers twice.

Use **`bindButton()`**, **`bindTextInput()`**, and related `bind*()` helpers — they dispatch through a single internal event channel.

---

## Text inputs: `useInput()` + `bindTextInput()`

```{r}
render = function(state, ns) {
  useInput("title")

  bindTextInput(
    ns, "title", "Title", state$title,
    placeholder = "Enter a title",
    update = "input"   # or "blur" (default)
  )
}
```

- `useInput()` registers a callback that writes event values into state
- `bindTextInput()` renders a text field wired to that callback

---

## Text input focus

Rebuilding text inputs on every keystroke destroys focus.

ShinyState **automatically splits** your UI when typing controls are present:

- **`bindTextInput()`**, **`bindTextArea()`**, and **`bindNumericInput()`** stay mounted in the shell
- **Everything else** (headings, tables, summaries, preview panels) live-updates on its own
- Works in **`tagList()`**, **`fluidPage()`**, **`fluidRow()` / `column()`** layouts

No `preview()` wrapper needed in normal apps.

```{r}
render = function(state, ns) {
  useInput("title")

  fluidPage(
    fluidRow(
      column(
        6,
        bindTextInput(ns, "title", "Title", state$title, update = "input"),
        bindTextArea(ns, "notes", "Notes", state$notes, update = "input")
      ),
      column(
        6,
        h3("Live preview"),
        p(state$title),
        p(state$notes)
      )
    )
  )
}
```

Use `update = "input"` for live preview while typing; `update = "blur"` (default) syncs on click-away.

---

## Optional: `preview()`

Use `preview()` only when you need a **manual** split inside a mixed subtree:

```{r}
render = function(state, ns) {
  useInput("title")

  tagList(
    bindTextInput(ns, "title", "Title", state$title, update = "input"),
    preview(
      h3(state$title),
      p("Characters:", nchar(state$title))
    )
  )
}
```

Most apps never need this — auto-split handles it.

---

## Other bound inputs

| Helper | Control |
|--------|---------|
| `bindTextArea()` | Multi-line text |
| `bindNumericInput()` | Number field |
| `bindCheckbox()` / `bindSwitch()` | Checkbox / toggle |
| `bindRadioButtons()` | Radio group |
| `bindCheckboxGroup()` | Multiple checkboxes |
| `bindSelect()` | Dropdown / multi-select |
| `bindSlider()` | Range slider |
| `bindDateInput()` | Date picker |

All pair with `useInput("field_id")` in `render()`.

---

## Inputs gallery demo

```{r}
shiny::runApp(
  system.file("examples/inputs-gallery", package = "ShinyState")
)
```

Two-column `fluidPage()` layout:

- Left: every `bind*()` control
- Right: live summary table + JSON state
- Title and notes use `update = "input"`; other fields update on change/blur

---

## Minimal counter app

```{r}
counter <- component(
  id = "counter",
  state = useState(count = 0L),
  render = function(state, ns) {
    useCallback("inc", function(s) s$set(count = s$count + 1L))
    useCallback("dec", function(s) s$set(count = s$count - 1L))

    tagList(
      h3(paste("Count:", state$count)),
      bindButton(ns, "dec", "−"),
      bindButton(ns, "inc", "+")
    )
  }
)
```

---

## Run the counter example

```{r}
shiny::runApp(
  system.file("examples/counter", package = "ShinyState")
)
```

Or from source before reinstalling:

```{r}
shiny::runApp("inst/examples/counter")
```

---

## Multi-page apps

Use Shiny tabs + one component per page. Give each tab an explicit `value` and use [serve_dormant()] so hidden tabs stay **dormant** (no render, effects, or events):

```{r}
shinyApp(
  ui = navbarPage(
    id = "pages",
    tabPanel("Dashboard", mount(counter_component), value = "dashboard"),
    tabPanel("Search", mount(filter_component), value = "search"),
    tabPanel("Reducer", mount(reducer_component), value = "reducer")
  ),
  server = function(input, output, session) {
    serve_dormant(
      session = session, input = input, output = output,
      navbar = "pages",
      dashboard = counter_component,
      search = filter_component,
      reducer = reducer_component
    )
  }
)
```

Component state is preserved when you switch tabs — only the visible page runs.

For a single always-visible component, use [serve()] as before.

---

## Multipage demo

```{r}
shiny::runApp(
  system.file("examples/multipage", package = "ShinyState")
)
```

Shows:

- Counter (`useState`, `effect`, `useMemo`)
- Filter + pagination (`effect` resets page)
- Task list (`useReducer`)

---

## Shared state across components

Components are isolated by default.

For cross-page state, use a small store with subscribers:

```{r}
store <- createStore(list(count = 0L, history = character()))

store$get()
store$set(count = 1L)
store$subscribe(function(data) { ... })
```

---

## Subscribing in an effect

```{r}
effect(
  deps = NULL,
  function(state) {
    unsubscribe <- shared_store$subscribe(function(data) {
      state$set(count = data$count, history = data$history)
    })

    state$set(count = shared_store$get()$count)
    unsubscribe  # cleanup on unmount
  }
)
```

---

## Shared-store demo

```{r}
shiny::runApp(
  system.file("examples/shared-store", package = "ShinyState")
)
```

- **Controls** page updates global count/history
- **History** page reflects changes live

---

## Under the hood

ShinyState still uses Shiny internally, but hides it:

1. Each component has a private state store
2. `render()` runs in a hook context
3. State changes bump an internal version counter
4. **Auto-split:** typing controls and other `bind*()` inputs stay in the stable shell
5. Display-only regions render in separate `uiOutput` slots that update live
6. While typing, only preview slots refresh — inputs keep focus
7. Effects run after render when deps change
8. Hidden tabs can stay **dormant** via [serve_dormant()] — no render, effects, or events until shown

Bootstrap / Shiny CSS from `fluidPage()` is preserved after auto-split.

---

## Mental model

| React | ShinyState |
|-------|------------|
| Component | `component()` |
| `useState` | `useState()` |
| `useEffect` | `effect()` / `useEffect()` |
| `useMemo` | `useMemo()` |
| `useReducer` | `useReducer()` |
| onClick handler | `useCallback()` + `bindButton()` |
| onChange handler | `useInput()` + `bindTextInput()` / `bind*()` |
| Live output region | Automatic (or optional `preview()`) |
| Tab lifecycle | `serve_dormant()` |

---

## Best practices

- Keep one concern per component
- Prefer `bindButton()` and `bind*()` helpers inside `render()` — not raw Shiny inputs
- Put summaries, tables, and charts anywhere in `render()` — auto-split keeps typing controls stable
- Use `update = "input"` on text fields when you want live preview while typing
- Use `effect()` for side effects, not UI logic
- Use `useMemo()` for expensive derived values
- Use `serve_dormant()` for tabbed apps so hidden pages stay dormant
- Mount multiple components with `mount()` / `serve()` / `serve_dormant()`

---

## Troubleshooting

| Symptom | Likely cause |
|---------|----------------|
| Blank component area | Old app process still running; restart R + app |
| Handler fires twice | Used `actionButton()` instead of `bindButton()` |
| Text input loses focus | Use `bindTextInput()` not raw `textInput()`; reinstall v0.4.0+ |
| Preview only updates on blur | Text field uses `update = "blur"`; try `update = "input"` |
| Layout looks unstyled | Reinstall v0.4.0+ (Bootstrap deps preserved after auto-split) |
| `bindButton` / `preview` not found | Reinstall latest package version |
| Hidden tab still runs effects | Use [serve_dormant()] with tab `value` ids |

---

## API cheat sheet

```{r}
component(id, ..., render)
useState(...)
effect(deps, fn)
useEffect(fn, deps)
useMemo(fn, deps)
useReducer(reducer, initial)
useCallback(input_id, fn)
useInput(input_id, state_field)
preview(...)   # optional; auto-split is usually enough
bindButton(ns, input_id, label)
bindTextInput(ns, input_id, label, value, update = "input")
bindSelect(ns, input_id, label, choices, selected)
mount(component)
serve(component, input, output, session)
serve_dormant(session, input, output, navbar, ...)
```

---

## Next steps

1. Run all four bundled examples
2. Convert one existing module to `component()`
3. Extract shared app state into a store pattern
4. Build a multi-tab dashboard from small components

---

## Thank you

Questions?

Package repo: `ShinyState`

Examples:

- `examples/counter`
- `examples/multipage`
- `examples/shared-store`
- `examples/inputs-gallery`
