---
title: "ShinyState Tutorial"
subtitle: "React-style components and hooks for Shiny"
author: "ShinyState"
date: "`r format(Sys.Date())`"
output:
  revealjs::revealjs_presentation:
    theme: moon
    transition: slide
    highlight: pygments
    center: false
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

## Welcome

**ShinyState** brings a React-inspired programming model to Shiny:

- Components with local state
- Hooks: `useState()`, `useEffect()`, `useMemo()`, `useReducer()`
- No hand-written `reactive()` / `observe()` in app code

---

## The problem

Classic Shiny spreads logic across:

- `reactive()` values
- `observe()` / `observeEvent()`
- `renderUI()` / `renderText()`

As apps grow, dependencies become hard to follow.

---

## The ShinyState idea

Think in **components**:

```r
component(
  id = "counter",
  state = useState(count = 0L),
  effect(deps = "count", function(state) { ... }),
  render = function(state, ns) {
    # hooks + UI
  }
)
```

Each component owns its state, effects, and render function.

---

## Install

```{r}
# install.packages("devtools")
devtools::install_local("path/to/ShinyState")

library(shiny)
library(ShinyState)
```

---

## Core workflow

1. Define components with `component()`
2. Mount UI with `mount()`
3. Run server logic with `serve()`

```{r}
shinyApp(
  ui = fluidPage(mount(counter)),
  server = function(input, output, session) {
    serve(counter, input, output, session)
  }
)
```

---

## `component()`

```{r}
counter <- component(
  id = "counter",
  state = useState(count = 0L),
  effect(
    deps = "count",
    function(state) {
      message("Count is now: ", state$count)
    }
  ),
  render = function(state, ns) {
  }
)
```

Returns a list with `ui`, `server`, and metadata.

---

## `useState()`

Declare initial state at component definition time:

```{r}
state = useState(
  page = 1L,
  filter = NULL
)
```

Inside `render()`, read and update via the `state` accessor:

```{r}
state$count          # read
state$set(count = 5) # write + re-render
state$all()          # snapshot
```

---

## State updates

Functional updates are supported:

```{r}
state$update(function(current) {
  list(count = current$count + 1L)
})
```

`state$set(...)` triggers a component re-render automatically.

---

## `effect()` / `useEffect()`

Run side effects when dependencies change.

**Declarative** (top-level in `component()`):

```{r}
effect(
  deps = "filter",
  function(state) {
    state$set(page = 1L)  # reset page when filter changes
  }
)
```

- `deps = NULL` → run once on mount
- Return a function from the effect to run cleanup

---

## `useMemo()`

Cache expensive derived values:

```{r}
render = function(state, ns) {
  doubled <- useMemo(
    function() state$count * 2L,
    deps = "count"
  )

  p(paste("Doubled:", doubled))
}
```

Recomputes only when `count` changes.

---

## `useReducer()`

For state transitions with actions:

```{r}
render = function(state, ns) {
  reducer <- useReducer(
    function(prev, action) {
      switch(action,
        add = c(prev, "new"),
        drop = if (length(prev)) prev[-length(prev)] else prev,
        prev
      )
    },
    initial = character()
  )

  reducer$dispatch("add")
  reducer$state
}
```

---

## `bindButton()` with `onClick`

Wire button clicks to state updates — the handler lives on the button:

```{r}
render = function(state) {
  bindButton("inc", "+", onClick = function(s) {
    s$set(count = s$count + 1L)
  })
}
```

**Important:** use `bindButton()`, not `actionButton()`, inside `render()`.

---

## Why bound helpers?

`render()` output is patched into the DOM when state changes.

Plain Shiny inputs and buttons can reset on re-render and fire handlers twice.

Use **`bindButton()`**, **`bindTextInput()`**, and related `bind*()` helpers —
they dispatch through a single internal event channel. Call them **inside
`render()`**: no `ns`, no separate `useInput()`, value read from state.

---

## Text inputs: `bindTextInput()`

```{r}
render = function(state) {
  bindTextInput(
    "title", "Title",
    placeholder = "Enter a title",
    update = "input"   # live (default); or "blur"
  )
}
```

- Reads its value from `state$title` and writes changes back automatically
- No `ns`, no `useInput()` — one line per field

---

## Text input focus

Rebuilding text inputs on every keystroke would destroy focus.

ShinyState renders into a single container that is **patched in place** (like a
virtual DOM): only changed nodes are touched, and the element you are currently
editing is never rebuilt.

- Type freely while summaries, tables, and previews update around you
- Cursor position, selection, and scroll all survive
- Works in `tagList()`, `fluidPage()`, `fluidRow()` / `column()` layouts

```{r}
render = function(state) {
  fluidPage(
    fluidRow(
      column(
        6,
        bindTextInput("title", "Title"),
        bindTextArea("notes", "Notes")
      ),
      column(
        6,
        h3("Live preview"),
        p(state$title),
        p(state$notes)
      )
    )
  )
}
```

Use `update = "input"` (default) for live updates while typing; `update = "blur"`
syncs on click-away.

---

## Other bound inputs

| Helper | Control |
|--------|---------|
| `bindTextArea()` | Multi-line text |
| `bindNumericInput()` | Number field |
| `bindCheckbox()` / `bindSwitch()` | Checkbox / toggle |
| `bindRadioButtons()` | Radio group |
| `bindCheckboxGroup()` | Multiple checkboxes |
| `bindSelect()` | Dropdown / multi-select |
| `bindSlider()` | Range slider |
| `bindDateInput()` | Date picker |

All pair with `useInput("field_id")` in `render()`.

---

## Inputs gallery demo

```{r}
shiny::runApp(
  system.file("examples/inputs-gallery", package = "ShinyState")
)
```

Two-column `fluidPage()` layout:

- Left: every `bind*()` control
- Right: live summary table + JSON state
- Title and notes use `update = "input"`; other fields update on change/blur

---

## Minimal counter app

```{r}
counter <- component(
  id = "counter",
  state = useState(count = 0L),
  render = function(state) {
    tagList(
      h3(paste("Count:", state$count)),
      bindButton("dec", "−", onClick = function(s) s$set(count = s$count - 1L)),
      bindButton("inc", "+", onClick = function(s) s$set(count = s$count + 1L))
    )
  }
)
```

---

## Run the counter example

```{r}
shiny::runApp(
  system.file("examples/counter", package = "ShinyState")
)
```

Or from source before reinstalling:

```{r}
shiny::runApp("inst/examples/counter")
```

---

## Multi-page apps

Pass **named** components to `shinyStateApp()`: one tab per component, the
dormant lifecycle, and URL routing, all wired for you. Names become page ids.

```{r}
shinyStateApp(
  title = "My App",
  routing = "hash",
  dashboard = counter_component,
  search = filter_component,
  reducer = reducer_component
)
```

Component state is preserved when you switch tabs — only the visible page runs.
Hidden tabs are **dormant** (no render, effects, or events). A single component
becomes a one-page app.

---

## URL routing

`routing = "hash"` gives every page a bookmarkable URL (`#!/dashboard`,
`#!/search`, ...). The browser back/forward buttons navigate between tabs:

- Component names double as page names in the URL
- Opening `https://myapp/#!/search` starts the app on the Search tab
- Unknown routes warn and fall back to the first component's page
- Link between pages with `route_link("search", "Go to search")` — a plain
  `<a href="#!/search">` anchor, no JavaScript needed
- Read the current route in any component with `useRoute()`

Under the hood this is [serve_dormant()] + [router_server()]; call them directly
if you build the `navbarPage` yourself.

---

## Multipage demo

```{r}
shiny::runApp(
  system.file("examples/multipage", package = "ShinyState")
)
```

Shows:

- Counter (`useState`, `effect`, `useMemo`)
- Filter + pagination (`effect` resets page)
- Task list (`useReducer`)

---

## Shared state across components

Components are isolated by default. For cross-page state, create a store and
use it from any component with `useStore()`:

```{r}
app_store <- createStore(count = 0L, history = character())

controls <- component(id = "controls", render = function(state) {
  store <- useStore(app_store)              # subscribes; auto-cleans up
  bindButton("inc", "+", onClick = function(s) {
    store$set(count = store$count + 1L)      # notifies every subscriber
  })
})
```

Every component using `useStore(app_store)` re-renders when the store changes.
No manual subscribe/unsubscribe, no effect plumbing.

---

## Shared-store demo

```{r}
shiny::runApp(
  system.file("examples/shared-store", package = "ShinyState")
)
```

- **Controls** page updates global count/history
- **History** page reflects changes live

---

## Under the hood

ShinyState still uses Shiny internally, but hides it:

1. Each component has a private state store
2. `render()` runs in a hook context and returns UI from state
3. State changes bump an internal version counter
4. **DOM patch:** the new UI is diffed against the live DOM; only changed nodes move
5. The element you're editing (focused input) is never rebuilt — focus and cursor survive
6. Effects run after render when deps change
7. Hidden tabs can stay **dormant** via [serve_dormant()] — no render, effects, or events until shown

---

## Mental model

| React / Vue | ShinyState |
|-------------|------------|
| Component | `component()` |
| `useState` | `useState()` |
| `useEffect` | `effect()` / `useEffect()` |
| computed | `computed = list(...)` |
| `watch` | `watch(fields, fn)` |
| `useMemo` | `useMemo()` |
| `useReducer` | `useReducer()` |
| onClick handler | `bindButton(onClick = ...)` |
| onChange / v-model | `bind*()` inputs |
| onMounted / onActivated | `onMounted()` / `onActivated()` |
| Pinia / Zustand store | `createStore()` / `useStore()` |
| props | `mount(child, props = ...)` |
| router | `routing = "hash"` + `useRoute()` |

---

## Best practices

- Keep one concern per component
- Call `bind*()` helpers and hooks inside `render()`, in the same order every render
- Put summaries, tables, and charts anywhere in `render()` — the patcher keeps inputs stable
- Use `computed` / `useMemo()` for derived and expensive values
- Use `effect()` / `watch()` for side effects, not UI logic
- Share cross-component state with `createStore()` / `useStore()`
- Build whole apps with `shinyStateApp()`

---

## Troubleshooting

| Symptom | Likely cause |
|---------|----------------|
| Blank component area | Old app process still running; restart R + app |
| Handler fires twice | Used `actionButton()` instead of `bindButton()` |
| `bind*()` errors "must be called inside render()" | Call bound inputs inside `render()`, not in the UI |
| Hook-order warning | A hook is called conditionally; call hooks every render |
| Prop not updating in a child | Parent must re-render (change parent state) to push props |
| Function not found | Reinstall latest package version |

---

## API cheat sheet

```{r}
component(id, ..., render, computed = list())
useState(...)                       # state
effect(deps, fn) / useEffect(fn, deps)
watch(fields, fn, immediate)        # fn(new, old)
useMemo(fn, deps)
useReducer(reducer, initial)
onMounted(fn) / onUnmounted(fn)
onActivated(fn) / onDeactivated(fn) # dormant tab wake/sleep

# bound inputs — inside render(), no ns
bindTextInput(input_id, label)
bindSelect(input_id, label, choices)
bindButton(input_id, label, onClick = function(s) ...)

# stores, routing, apps
createStore(...) / useStore(store)
useRoute() / route_link(page, label)
mount(component, props = list())
shinyStateApp(..., title, routing = "hash")
```

---

## Next steps

1. Run the bundled examples (counter, multipage, shared-store, children, inputs-gallery)
2. Convert one existing module to `component()`
3. Extract shared app state into a `createStore()`
4. Build a multi-tab dashboard with `shinyStateApp()`

---

## Thank you

Questions?

Package repo: `ShinyState`

Examples:

- `examples/counter`
- `examples/multipage`
- `examples/shared-store`
- `examples/children`
- `examples/inputs-gallery`
