---
title: "ShinyState Tutorial"
subtitle: "React-style components and hooks for Shiny"
author: "ShinyState"
date: "`r format(Sys.Date())`"
output:
  revealjs::revealjs_presentation:
    theme: moon
    transition: slide
    highlight: pygments
    center: false
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

## Welcome

**ShinyState** brings a React-inspired programming model to Shiny:

- Components with local state
- Hooks: `useState()`, `useEffect()`, `useMemo()`, `useReducer()`
- No hand-written `reactive()` / `observe()` in app code

---

## The problem

Classic Shiny spreads logic across:

- `reactive()` values
- `observe()` / `observeEvent()`
- `renderUI()` / `renderText()`

As apps grow, dependencies become hard to follow.

---

## The ShinyState idea

Think in **components**:

```r
component(
  id = "counter",
  state = useState(count = 0L),
  effect(deps = "count", function(state) { ... }),
  render = function(state, ns) {
    # hooks + UI
  }
)
```

Each component owns its state, effects, and render function.

---

## Install

```{r}
# install.packages("devtools")
devtools::install_local("path/to/ShinyState")

library(shiny)
library(ShinyState)
```

---

## Core workflow

1. Define components with `component()`
2. Mount UI with `mount()`
3. Run server logic with `serve()`

```{r}
shinyApp(
  ui = fluidPage(mount(counter)),
  server = function(input, output, session) {
    serve(counter, input, output, session)
  }
)
```

---

## `component()`

```{r}
counter <- component(
  id = "counter",
  state = useState(count = 0L),
  effect(
    deps = "count",
    function(state) {
      message("Count is now: ", state$count)
    }
  ),
  render = function(state, ns) {
  }
)
```

Returns a list with `ui`, `server`, and metadata.

---

## `useState()`

Declare initial state at component definition time:

```{r}
state = useState(
  page = 1L,
  filter = NULL
)
```

Inside `render()`, read and update via the `state` accessor:

```{r}
state$count          # read
state$set(count = 5) # write + re-render
state$all()          # snapshot
```

---

## State updates

Functional updates are supported:

```{r}
state$update(function(current) {
  list(count = current$count + 1L)
})
```

`state$set(...)` triggers a component re-render automatically.

---

## `effect()` / `useEffect()`

Run side effects when dependencies change.

**Declarative** (top-level in `component()`):

```{r}
effect(
  deps = "filter",
  function(state) {
    state$set(page = 1L)  # reset page when filter changes
  }
)
```

- `deps = NULL` → run once on mount
- Return a function from the effect to run cleanup

---

## `useMemo()`

Cache expensive derived values:

```{r}
render = function(state, ns) {
  doubled <- useMemo(
    function() state$count * 2L,
    deps = "count"
  )

  p(paste("Doubled:", doubled))
}
```

Recomputes only when `count` changes.

---

## `useReducer()`

For state transitions with actions:

```{r}
render = function(state, ns) {
  reducer <- useReducer(
    function(prev, action) {
      switch(action,
        add = c(prev, "new"),
        drop = if (length(prev)) prev[-length(prev)] else prev,
        prev
      )
    },
    initial = character()
  )

  reducer$dispatch("add")
  reducer$state
}
```

---

## `useCallback()` + `bindButton()`

Wire button clicks to state updates:

```{r}
render = function(state, ns) {
  useCallback("inc", function(s) {
    s$set(count = s$count + 1L)
  })

  bindButton(ns, "inc", "+")
}
```

**Important:** use `bindButton()`, not `actionButton()`, inside `render()`.

---

## Why `bindButton()`?

`render()` output is rebuilt when state changes.

Plain `actionButton()` values reset on re-render and can fire handlers twice.

`bindButton()` dispatches through a single internal event channel.

---

## Minimal counter app

```{r}
counter <- component(
  id = "counter",
  state = useState(count = 0L),
  render = function(state, ns) {
    useCallback("inc", function(s) s$set(count = s$count + 1L))
    useCallback("dec", function(s) s$set(count = s$count - 1L))

    tagList(
      h3(paste("Count:", state$count)),
      bindButton(ns, "dec", "−"),
      bindButton(ns, "inc", "+")
    )
  }
)
```

---

## Run the counter example

```{r}
shiny::runApp(
  system.file("examples/counter", package = "ShinyState")
)
```

Or from source before reinstalling:

```{r}
shiny::runApp("inst/examples/counter")
```

---

## Multi-page apps

Use Shiny tabs + one component per page:

```{r}
shinyApp(
  ui = navbarPage(
    "Multipage",
    tabPanel("Dashboard", mount(counter_component)),
    tabPanel("Search", mount(filter_component)),
    tabPanel("Reducer", mount(reducer_component))
  ),
  server = function(input, output, session) {
    serve(counter_component, input, output, session)
    serve(filter_component, input, output, session)
    serve(reducer_component, input, output, session)
  }
)
```

---

## Multipage demo

```{r}
shiny::runApp(
  system.file("examples/multipage", package = "ShinyState")
)
```

Shows:

- Counter (`useState`, `effect`, `useMemo`)
- Filter + pagination (`effect` resets page)
- Task list (`useReducer`)

---

## Shared state across components

Components are isolated by default.

For cross-page state, use a small store with subscribers:

```{r}
store <- createStore(list(count = 0L, history = character()))

store$get()
store$set(count = 1L)
store$subscribe(function(data) { ... })
```

---

## Subscribing in an effect

```{r}
effect(
  deps = NULL,
  function(state) {
    unsubscribe <- shared_store$subscribe(function(data) {
      state$set(count = data$count, history = data$history)
    })

    state$set(count = shared_store$get()$count)
    unsubscribe  # cleanup on unmount
  }
)
```

---

## Shared-store demo

```{r}
shiny::runApp(
  system.file("examples/shared-store", package = "ShinyState")
)
```

- **Controls** page updates global count/history
- **History** page reflects changes live

---

## Under the hood

ShinyState still uses Shiny internally, but hides it:

1. Each component has a private state store
2. `render()` runs in a hook context
3. State changes invalidate an internal counter
4. `renderUI()` rebuilds component UI
5. Effects run after render when deps change

---

## Mental model

| React | ShinyState |
|-------|------------|
| Component | `component()` |
| `useState` | `useState()` |
| `useEffect` | `effect()` / `useEffect()` |
| `useMemo` | `useMemo()` |
| `useReducer` | `useReducer()` |
| onClick handler | `useCallback()` + `bindButton()` |

---

## Best practices

- Keep one concern per component
- Prefer `bindButton()` for actions inside `render()`
- Use `effect()` for side effects, not UI logic
- Use `useMemo()` for expensive derived values
- Mount multiple components with `mount()` / `serve()`

---

## Troubleshooting

| Symptom | Likely cause |
|---------|----------------|
| Blank component area | Old app process still running; restart R + app |
| Handler fires twice | Used `actionButton()` instead of `bindButton()` |
| `bindButton` not found | Reinstall latest package version |
| Example path is `""` | Example not in installed package yet |

---

## API cheat sheet

```{r}
component(id, ..., render)
useState(...)
effect(deps, fn)
useEffect(fn, deps)
useMemo(fn, deps)
useReducer(reducer, initial)
useCallback(input_id, fn)
bindButton(ns, input_id, label)
mount(component)
serve(component, input, output, session)
```

---

## Next steps

1. Run all three bundled examples
2. Convert one existing module to `component()`
3. Extract shared app state into a store pattern
4. Build a multi-tab dashboard from small components

---

## Thank you

Questions?

Package repo: `ShinyState`

Examples:

- `examples/counter`
- `examples/multipage`
- `examples/shared-store`
